\chapter*{Related Research Material}
\addcontentsline{toc}{chapter}{Related Research Material}

This PhD thesis has led to the development of following materials:\\

\textbf{Conference papers}
\begin{itemize}
\item \textit{Qualitative {P}ossibilistic {M}ixed-{O}bservable {MDP}s}, 
Proceedings of the Twenty-Ninth Annual Conference on Uncertainty in Artificial Intelligence (UAI-13) 
with Florent Teichteil-Konigsbuch, Jean-Loup Farges, and Didier Dubois, (2013) \cite{Drougard13}. Presentation: poster and 2-minute speech in English.
\item \textit{Structured Possibilistic Planning Using Decision Diagrams}, 
Proceedings of the Twenty-Eighth {AAAI} Conference on Artificial Intelligence,
(AAAI-14) with Florent Teichteil{-}K{\"{o}}nigsbuch, Jean-Loup Farges and Didier Dubois,
(2014) \cite{DBLP:conf/aaai/DrougardTFD14}. Presentation: poster and 20-minute speech in English.
\item \textit{Planning in Partially Observable Domains with Fuzzy Epistemic States and Probabilistic Dynamics},
Proceedings of the Ninth International Conference on Scalable Uncertainty Management (SUM-15)
with Didier Dubois, Jean-Loup Farges and Florent Teichteil-K\"onigsbuch, (2015) \cite{DBLP:conf/sum/DrougardDFT15}. Presentation: 30-minute speech in English.
\item \textit{Processus D\'ecisionnels de Markov Possibilistes \`a Observabilit\'e Mixte},
Eighth Journ\'ees Francophones sur la Planification, la D\'ecision et l'Apprentissage pour la conduite de syst\`emes (JFPDA-13),
with Florent Teichteil-Konigsbuch, Jean-Loup Farges, and Didier Dubois, (2013). Presentation: poster and 30-minute speech in French.
\item \textit{Planification dans des domaines partiellement
observables avec des \'etats \'epist\'emiques flous
et une dynamique probabiliste.},
Tenth Journ\'ees Francophones sur la Planification, la D\'ecision et l'Apprentissage pour la conduite de syst\`emes (JFPDA-15),
with Didier Dubois, Jean-Loup Farges and Florent Teichteil-K\"onigsbuch, (2015). Presentation: 30-minute speech in French.
\end{itemize}

\textbf{Workshop paper}
\begin{itemize}
\item \textit{Structured Possibilistic Planning Using Decision Diagrams}, 
Workshop ``Models and Paradigms for Planning under Uncertainty: a Broad Perspective'' 
of the Twenty-Fourth International Conference on Automated Planning and Scheduling (ICAPS-14).
Presentation by Florent Teichteil-K\"onigsbuch.
\end{itemize}

\textbf{Journal papers}
\begin{itemize}
\item \textit{A Possibilistic Estimation of Human Attentional Errors},
submitted to the IEEE Transactions on Fuzzy Systems (TFS),
with Sergio Pizziol, Catherine Tessier, Jean-Loup Farges, Didier Dubois and Fr\'ed\'eric Dehais.
\item \textit{Processus D\'ecisionnels de Markov Possibilistes \`a Observabilit\'e Mixte},
Special edition of the Revue d'Intelligence Artificielle (RIA. Volume 29 -- n6/2015),
with Florent Teichteil-Konigsbuch, Jean-Loup Farges, and Didier Dubois.
\end{itemize}

\textbf{Oral presentations}
\begin{itemize}
\item ``Possibilistic MDP and POMDP''. Seminar in English of the SequeL team at INRIA Lille, France (2012). 
\item ``Mod\`eles possibilistes pour la d\'ecision s\'equentielle dans l'incertain''. Seminar in French of the MAD team at GREYC Caen, France (2013).
\item EDSYS days, doctoral school conference: 30-minute speech in French at ISAE-SUPAERO, Toulouse France (2014). 
\item PhD student days (JDD), 20-minute speech in French at Onera -- The French Aerospace Lab, Toulouse France (2014 and 2015).
\item Onera-DLR Aerospace Symposium, (ODAS 2015), 30-minute speech in English, at Onera -- The French Aerospace Lab, Toulouse France (2015).
\end{itemize}

\textbf{Code}
\begin{itemize}
\item Participation in the International Probabilistic Planning Competition 2014 -- MDP track, with the algorithm PPUDD.
\item Github repository \url{http://www.github.com/drougui/ppudd}.
\end{itemize}
